---
title: "GY476_OSMR"
author: "Elisabetta Pietrostefani & Louise Bernard"
date: "21/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(jsonlite)
library(osmdata)
library(osrm)
library(osmdata)
library(sf)
library(dplyr)
library(tmap)
```

**Task 1** - Using the the OSMrouting service calculate travel between the LSE and the different bike stations. 
Refer back to the OSM R-markdown. What type of network analysis is this?

Point-to-Point analysis - trvael time

``` {r distancetolse}
# Load the hire bike locations from the London API

# https://www.any-api.com/tfl_gov_uk/tfl_gov_uk/console/_BikePoint/BikePoint_GetAll
result <- fromJSON("../GY476_data_2022_23/London/NetworkAnalysis/BikePoint.json")

# create new ID
result$`Station ID` = as.numeric(substr(result$id, nchar("BikePoints_")+1, nchar(result$id)))

# Print the column names
names(result)

# create a sf object
stations_df <- result %>% 
  sf::st_as_sf(coords = c(10,9))  %>%  # create pts from coordinates
  st_set_crs(4326) %>%  # set the original CRS
  relocate(`Station ID`) # set ID as the first column of the dataframe
```

Find LSE coordinates
``` {r lse coordinates}
#find LSE coordinates
q <- opq("greater london uk") %>%
    add_osm_feature(key = "name", value = "London School of Economics and Political Science") %>%
  osmdata_sf()

# get the centroid of the polygons
lse_centre <- q$osm_multipolygons %>% st_centroid()

```

Travel time matrix with different sets of origins and destinations.
There is a **maximum size request**, so we are going to split our request into smaller ones

**OSRM supports "profiles"** 
Profiles representing routing behavior for different transport modes like car, bike and foot. 
You can also create profiles for variations like a fastest/shortest car profile or fastest/safest/greenest bicycles profile.
https://github.com/Project-OSRM/osrm-backend/blob/master/docs/profiles.md

``` {r lse coordinates}
# list all the stations
N_stations <- 1:dim(stations_df)[1]

# divide in 100 stations chuncks
chuncks <- split(N_stations, ceiling(seq_along(N_stations)/100))

# create a function that request the distances to lse and wait 0.05 seconds
DistLSE <- function(i) {
  Sys.sleep(0.05)
  distA <- osrmTable(src = lse_centre, dst = stations_df[i,], osrm.profile = "bike") 
  df <- data.frame(distA$destinations, t(distA$durations))
  names(df) <- c("lat", "lon", "durations")
  return(df)
}
# src is a vector of identifier, longitude and latitude (WGS84) of origin point
# dst is a vector of identifier, longitude and latitude (WGS84) of desintation point
# osrm.profile is bike!
```

Get distance to lse
``` {r distances}
distance_to_lse <- lapply(chuncks, DistLSE)
distance_to_lse <- do.call("rbind", distance_to_lse)

# attach back to the original dataframe
stations_df <- cbind(stations_df, distance_to_lse)

# have a look at distance_to_lse
# and stations_df
```

What have you done?

# add a background map for more context
``` {r Background map}
tmap_mode("view")
tm_basemap() +
  tm_shape(stations_df) +
  tm_symbols(id = "commonName", col = "durations", scale = .5)
```

**Task 2** - What type of Network Analysis have you conducted here? What does is show?

The most population trips are made around Hype Park, Westfield shopping centre, King's Cross and Waterloo stations.
