---
title: "GY476"
subtitle: "Revision Session 2022/23"
author: "Elisabetta Pietrostefani" 
date: "6/12/2022"
output:
  html_document:
    df_print: paged
---

# R Markdown Formatting

https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

## Display Options - do not display outputs that are not requested in the assignment

For example in the Formative Assessment some of you displayed density rasters of the world.

I created four Rmarkdown code chuncks below to explain how to control the rmarkdown output.

The first chunck uses Rmarkdown standard option: the code will be evaluated (`eval` = `TRUE`) and it will be displayed with the results (`echo` = `TRUE`).

In the second chunck, I added the option `include` = `FALSE`. The code and its output will not be displayed. It will run normally.

Finally, in the third chunk, I set the default option to no longer display the code. You will notice that from now on, the code is no longer displayed, but any output still is.

```{r}
# create a sum and display results
1+1
```

```{r, include = FALSE}
# create the sum of 1+1 but do not display the code or the result
s <- 1+1
```

```{r}
knitr::opts_chunk$set(echo = FALSE) # do not display code for all following chuncks
```

```{r}
# display result only
s
```

# Let's revise improving maps

```{r}
knitr::opts_chunk$set(echo = TRUE) # display codes in the final document

library(sf)
wards <- read_sf("../GY476_data_2022_23/London/Polygons/wards.shp")

# the simplest map 
plot(wards$geometry)

# add some grey fill to the wards
plot(wards$geometry, col="grey")

# add a title
plot(wards$geometry, col="grey", main = "London wards")

# plot a variable
wards$level4p <- wards$LEVEL4_5/wards$POP16_74
plot(wards[,"level4p"], main = "Level 4 and over")

# plot quantiles
plot(wards[,"level4p"], main = "Level 4 and over", breaks = "quantile")

# select 1st quartile, median, 3rd quartile, and max
summary(wards[,"level4p"])
plot(wards[,"level4p"], main = "Level 4 and over", breaks = "quantile", nbreaks = 4)

# choose a new colour palette and set higher values to be red
summary(wards[,"level4p"])
plot(wards[,"level4p"], main = "Level 4 and over", breaks = "quantile", nbreaks = 4, pal = heat.colors(4, rev = TRUE))

# if you want to change how the variable displays - change the variable ;)
wards$level4pp <- wards$LEVEL4_5/wards$POP16_74*100
summary(wards[,"level4pp"])
plot(wards[,"level4pp"], main = "Level 4 and over", breaks = "quantile", nbreaks = 4, pal = heat.colors(4, rev = TRUE))

```

The command `plot` does not allowed to add additional data, improve the legend, or add a basemap easily. `ggplot2` offers more graphics options:

```{r}
library(ggplot2)

# simple map using ggplot2 and sf
ggplot(data = wards) +
    geom_sf()
```

```{r}
# If I wanted to add labels to my x and y axis
ggplot(data = wards) +
    geom_sf() +
labs(x=expression(paste("Longitude ",degree,"W",sep="")),
       y=expression(paste("Latitude ",degree,"N",sep="")))
```

```{r}
# remove background
ggplot() +
  geom_sf(data = wards) +
  theme(line = element_blank(), # remove tick marks
        rect = element_blank(),  # remove background
        axis.text=element_blank()) # remove x and y axis  
```

```{r}
        
# add share of people with level 4 education 
ggplot() +
  geom_sf(data = wards, aes(fill = level4p)) +
  theme(line = element_blank(), # remove tick marks
        rect = element_blank(), # remove background
        axis.text=element_blank()) # remove x and y axis
```

```{r}

# reverse the colour to make higher values darker 
ggplot() +
  geom_sf(data = wards, aes(fill = -level4p)) +
  theme(line = element_blank(), # remove tick marks
        rect = element_blank(), # remove background
        axis.text=element_blank()) # remove x and y axis

```

```{r}
# change the colour to a colour gradient with the mean as midpoint
ggplot() +
  geom_sf(data = wards, aes(fill = level4p)) +
  theme(line = element_blank(), # remove tick marks
        rect = element_blank(), # remove background
        axis.text=element_blank()) + # remove x and y axis
  scale_fill_gradient2("Level 4 and over", 
                       low = "yellow", 
                       mid="white", 
                       high = "red", 
                       midpoint=mean(wards$level4p))

```

```{r}

# add a title
ggplot() +
  geom_sf(data = wards, aes(fill = level4p)) +
  theme(line = element_blank(), # remove tick marks
        rect = element_blank(), # remove background
        axis.text=element_blank()) + # remove x and y axis
  scale_fill_gradient2("Level 4 and over", # legend title
                       low = "yellow", # first colour
                       mid="white", # midpoint
                       high = "red", # high
                       midpoint=mean(wards$level4p)) + # set mid point to mean
  ggtitle("Share of population with level 4 education") # add title

```

```{r}
# add a scale bar 
library(ggspatial)
ggplot() +
  geom_sf(data = wards, aes(fill = level4p)) +
  theme(line = element_blank(), # remove tick marks
        rect = element_blank(), # remove background
        axis.text=element_blank()) + # remove x and y axis
  scale_fill_gradient2("Level 4 and over", # legend title
                       low = "yellow", # first colour
                       mid="white", # midpoint
                       high = "red", # high
                       midpoint=mean(wards$level4p)) + # set mid point to mean
  ggtitle("Share of population with level 4 education") + # add title
  annotation_scale() + # add a scale
  annotation_north_arrow(pad_y = unit(1, "cm")) # add an arrow and move it up by 1 cm

```

```{r}
# create quantiles using classIntervals
library(classInt)
library(dplyr) 

# find the quantiles breaks
breaks_qt <- classIntervals(wards$level4p, n = 4, style = "quantile")

# create quantiles using the command cut
# https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/cut
wards <- mutate(wards, london_edu_cat = cut(level4p, breaks_qt$brks)) 

# plot
ggplot() +
  geom_sf(data = wards, aes(fill = london_edu_cat)) +
  theme(line = element_blank(), # remove tick marks
        rect = element_blank(), # remove background
        axis.text=element_blank()) + # remove x and y axis
  scale_fill_brewer(palette = "OrRd") +
  ggtitle("Share of population with level 4 education") +
  annotation_scale() + # add a scale
  annotation_north_arrow(pad_y = unit(1, "cm")) # add a arrow and move it up by 1 cm
```

```{r}
# there is one missing value and too many digits in our legend
# we can investigate quickly where the missing value comes from
summary(wards)
head(wards[is.na(wards$london_edu_cat),])

# it comes from the lowest value not being included

```

```{r}
# we fix the problems by changing how we create our quartiles:
# include the lowest values of our intervals and only one digit to the labels
wards <- mutate(wards,
                london_edu_cat = cut(level4p, breaks_qt$brks, include.lowest = TRUE, dig.lab	=1)) 

# plot
ggplot() +
  geom_sf(data = wards, aes(fill = london_edu_cat)) +
  theme(line = element_blank(), # remove tick marks
        rect = element_blank(), # remove background
        axis.text=element_blank()) + # remove x and y axis
  scale_fill_brewer("Share level 4", palette = "OrRd") +
  ggtitle("Share of population with level 4 education") +
  annotation_scale() + # add a scale
  annotation_north_arrow(pad_y = unit(1, "cm")) # add a arrow and move it up by 1 cm


```

```{r}
# I am still not happy with my legend, I would like nice percentage
# plot
library(Hmisc)
library(scales)
wards <- mutate(wards,
                london_edu_cat2 = Hmisc::cut2(wards$level4p, 
                                              breaks_qt$brks, 
                                              formatfun = scales::percent,
                                              digits=1)) 

# final plot
ggplot() +
  geom_sf(data = wards, aes(fill = london_edu_cat2)) +
  theme(line = element_blank(), # remove tick marks
        rect = element_blank(), # remove background
        axis.text=element_blank()) + # remove x and y axis
  scale_fill_brewer("Share level 4", palette = "OrRd") +
  ggtitle("Share of population with level 4 education") +
  annotation_scale() + # add a scale
  annotation_north_arrow(pad_y = unit(1, "cm")) # add a arrow and move it up by 1 cm


```

Interactive maps are also a great way to visualize data:

```{r map1}
library(tmap)
# a basic map of the share of people with level 4 education in London
tmap_mode("view")
tm_basemap() +
  tm_shape(wards) +
  tm_polygons("level4p", palette = "YlOrBr") 
```

```{r map2}
# change the id to ward name
tm_basemap() +
  tm_shape(wards) +
  tm_polygons("level4p", palette = "YlOrBr", id = "WARD_NAME") 
```

```{r map3}
# map quartiles
tm_basemap() +
  tm_shape(wards) +
  tm_polygons("level4p", palette = "YlOrBr", id = "WARD_NAME", n = 4, style= "quantile")   
```

```{r map4}
# change back to the nice cuts
tm_basemap() +
  tm_shape(wards) +
  tm_polygons("london_edu_cat2", palette = "YlOrBr", id = "WARD_NAME",  title = "Share level 4")
```

```{r}
tmap_mode("plot")
# let's plot the population
tm_shape(wards) +
  tm_fill("POP16_74",style="fixed", title = "Population", breaks=c(0, 4000, 6000, 8000, 10000, 16000))+
  tm_borders() +
  tm_layout(legend.outside = TRUE, legend.outside.position = "right") 

# it is not great as larger wards have a higher population 
# let's normalize by area
wards <- wards %>% 
  mutate(area = as.numeric(st_area(.)),
         `Population density` = POP16_74/area)


# summarize the new variable
summary(wards)

# the min is 0.0001 and max 0.01, not quite easy to interpret
# area is calculated in m2, but people per km2 would be much easier to visualize
# let's replace area in m2 by km2
wards <- wards %>%
  mutate(area = as.numeric(st_area(.)/1000000),# convert to km2
         `Population density` = POP16_74/area)

# and now plot the population density
tm_shape(wards) +
  tm_fill("Population density",
          style="fixed",
          breaks=c(0, 100, 2000, 5000, 8000, 18000),
          title = "Population density")+
  tm_borders() +
  tm_layout(legend.outside = TRUE, legend.outside.position = "right") 

# add a title, scale bar and arrow
tm_shape(wards) +
  tm_fill("Population density",
          style="fixed",
          breaks=c(0, 100, 2000, 5000, 8000, 18000),
          title = "Population density")+
  tm_borders() +
  tm_layout(title= "Map of London",
            legend.outside = TRUE, legend.outside.position = "right") +
  tm_scale_bar(position=c("left", "bottom"))+
  tm_compass()
  
```

# Let's Revise Scatter plot

```{r scatter1}
# Basic scatter plot
ggplot(wards, aes(x=`Population density`, y=level4p)) + 
  geom_point()
```

```{r scatter 2}
# Change axes labels
ggplot(wards, aes(x=`Population density`, y=level4p)) + 
  geom_point()+
  labs(y = "Share level 4")
```

```{r scatter 3}
# Change the axe scales 
ggplot(wards, aes(x=`Population density`, y=level4p)) + 
  geom_point()+
  labs(y = "Share level 4") + 
  scale_y_continuous(labels = scales::percent_format(scale = 100)) +
  scale_x_continuous(labels=comma)
```

```{r scatter 4}
# add a regression line
ggplot(wards, aes(x=`Population density`, y=level4p)) + 
  geom_point()+
  labs(y = "Share level 4") + 
  scale_y_continuous(labels = scales::percent_format(scale = 100)) +
  scale_x_continuous(labels=comma)  + 
  geom_smooth(method='lm', formula= y~x)
```

```{r scatter 5}
# add the formula
library(ggpmisc)

ggplot(wards, aes(x=`Population density`, y=level4p)) + 
  geom_point()+
  labs(y = "Share level 4") + 
  scale_y_continuous(labels = scales::percent_format(scale = 100)) +
  scale_x_continuous(labels=comma)  + 
  geom_smooth(method='lm', formula= y~x) + 
  stat_poly_eq(formula = y ~ x, 
                eq.with.lhs = "italic(hat(y))~`=`~",
                aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")), 
                parse = TRUE)


```

# Correlation matrices

```{r cormatrix}
library(corrplot)
wards_levels <- wards %>% st_drop_geometry() %>% select(starts_with("LEVEL"))

pairs(wards_levels)
correlations <- cor(wards_levels)
corrplot(correlations)

```

# Dropping outliers and dplyr pipes

```{r}
# let's look at airbnbs in Buenos Aires
BA <- read_sf("../GY476_data_2022_23/InsideAirbnb/neighbourhoods_BA_posgar.shp")

# read the AirBnb listing
listings <- read.csv("../GY476_data_2022_23/InsideAirbnb/listings_nooutliers.csv")
summary(listings$price)
hist(listings$price)

# locate the longitude and latitude
names(listings)

listings_clean = listings %>% # select the main object
  filter(price < 30000) %>% # remove values above 500,000
  st_as_sf(coords = c(8,7)) %>% # create sf
  st_set_crs(4326) %>% # set crs
  st_transform(st_crs(BA))

# plot
plot(BA$geometry)
plot(listings_clean$geometry, add=TRUE)

hist(listings_clean$price)
```

## Buffer calculations

Let's calculate the average population around a sample of listed buildinds. First, we need some population data. Then, we can download the census Output Areas for 2011.

```{r}
# load building
listed_buildings <- read_sf("../GY476_data_2022_23/London/Points/listed_buildings_london.shp")

# select randomly 20 ids
sample_ids <- sample(listed_buildings$ListEntry, 20)

# filter to the randomely selected 20 buildings
listed_buildings <- listed_buildings[listed_buildings$ListEntry %in% sample_ids,]

```

```{r donwloadoas}
# create output folder
dir.create("GY476_data_2022_23")
download.file("https://data.london.gov.uk/download/statistical-gis-boundary-files-london/9ba8c833-6370-4b11-abdc-314aa020d5e0/statistical-gis-boundaries-london.zip", destfile = "../GY476_Files_2021_22/London/statistical-gis-boundaries-london.zip")
# unzip
unzip("../GY476_data_2022_23/London/statistical-gis-boundaries-london.zip", exdir = "../GY476_data_2022_23/London")

# load 
oas_census <- read_sf("../GY476_data_2022_23/London/statistical-gis-boundaries-london/ESRI/OA_2011_London_gen_MHW.shp")

# calculate areas
oas_census$area_census <- st_area(oas_census)

# CHECK OUT what you created
oas_census

# plot
plot(oas_census$geometry)
```

The next step is to create a buffer around the buildings:

```{r buffer}
# project the data
listed_buildings <- st_transform(listed_buildings, st_crs(oas_census))
# create 150m buffer
buffers500m <- st_buffer(listed_buildings, dist=500)
# attach distance
buffers500m$area_buffer <- st_area(buffers500m)
```

```{r plot}
#plotting listed buildings and buffers
ggplot() +  
  geom_sf(data = listed_buildings$geometry, inherit.aes = FALSE, colour="#00abff")+
  geom_sf(data = buffers500m$geometry, fill=NA) 
```

And calculate the intersection of the Output Areas.

```{r inter, message=FALSE}
# create intersection of all the oas within the buffers
oas_census <- st_make_valid(oas_census) #can add this in if your R says there is a precision issue
buffers_inter <- st_intersection(buffers500m, oas_census)
buffers_inter$area_inter <- st_area(buffers_inter)

```

Finally, we can aggregate up at the building level to find the average population living within 500m of a listed building:

```{r aggregate}
# aggregate at the buffer level
buffers_agg <- buffers_inter %>% 
  mutate(weight = as.numeric(area_inter/area_buffer),
         population = as.numeric(USUALRES)) %>%
  group_by(ListEntry) %>%
  summarise_at("population", list(sum = ~sum(. * weight)))

round(mean(buffers_agg$sum))

```

The average population is `r round(mean(buffers_agg$sum))`!

# Interpolation

```{r krig}
library(raster)
library(gstat)

# Create geostatistical prediction using a ordinary Kriging
gs <- gstat(formula = price~1, # formula that defines the dependent variable as a linear model of independent variables
            data = listings_clean, # housesales data 
            nmax = 500, # the number of nearest observations that should be used
            set=list(idp = 0.2)) # set inverse distance power  

# Create a blank raster using district extent and crs
r <- raster(BA, resolution=200) # output cell size 

# Create the interpolated raster using the prediction from the model and the blank raster
idw <- interpolate(r, gs)

# Mask values outside London
idwr <- mask(idw, BA)

# Plot result
plot(idwr, col = rev(heat.colors(100)))
plot(BA$geometry, fill = NA, add=T)

```

# Let's revise Geographic Coordinate Systems vs. Projected Coordinate systems.

It is **good practice** to re-project your data for both for measurement and for lack of distortion.

```{r}
library("osrm")
library("osmdata")
library("dplyr")

q <- opq("greater london uk") %>%
    add_osm_feature(key = "name", value = "London School of Economics and Political Science") %>%
  osmdata_sf()
plot(q$osm_multipolygons$geometry)

lse_centre <- q$osm_multipolygons %>% st_centroid()
plot(lse_centre$g, add=TRUE)

crs(lse_centre)
st_crs(lse_centre)$units_gdal # to find out the CRS units
st_crs(lse_centre)$srid # extracts its SRID (when available)

lse_centre_reproj <- st_transform(lse_centre, crs = 27700)
crs(lse_centre_reproj)
st_crs(lse_centre_reproj)$units_gdal

```

`st_set_crs`adds a CRS to a sf object but it does not transform

You can get away using `st_distance` and `st_area` to calculate great circle distance by default in R. Measurement error is not that bad if you forget to project the data. However, I do not recommend this!

# References

## Maps in R

https://cengel.github.io/R-spatial/mapping.html

https://r-spatial.org/r/2018/10/25/ggplot2-sf.html

https://ggplot2.tidyverse.org/reference/ggsf.html

## ggplot2

http://www.sthda.com/english/wiki/ggplot2-scatter-plots-quick-start-guide-r-software-and-data-visualization

## dplyr

https://seananderson.ca/2014/09/13/dplyr-intro/
