
# **Simple Features**

## Basic Map making

As the first vignette explains, **simple feature objects in R** are stored in a data frame, with geographic data occupying a special column, usually named 'geom' or 'geometry.'

```{r}
vignette(package = "sf") # see which vignettes are available
vignette("sf1")          # an introduction to the package
```

We will use the world dataset provided by the spData. Note that instead of creating a single map, as most GIS programs would, the plot() command has created multiple maps, one for each variable in the world datasets. This behavior can be useful for exploring the spatial distribution of different variables.

```{r}
library(spData)
names(world)
plot(world)
summary(world["gdpPercap"])
```

Basic maps are created in sf with plot(). A legend or 'key' with a continuous color is produced if the object to be plotted has a single variable (see the right-hand panel). Colors can also be set with col =

```{r}
plot(world[3:6])
plot(world["gdpPercap"])
```

And we can also use ggplot2

```{r}
library(ggplot2)
ggplot() + geom_sf(data = world, aes(fill = gdpPercap))
```

The subsequent code chunk combines countries in Africa:

```{r}
world_africa = world[world$continent == "Africa", ]
africa = st_union(world_africa)
```

We can now plot the African continent over a map of the world. Note that the first plot must only have one facet for add = TRUE to work. If the first plot has a key, reset = FALSE must be used (result not shown):

```{r}
plot(world["gdpPercap"], reset = FALSE)
plot(africa, add = TRUE, col = "red")
```

sf's plot method also has arguments specific to geographic data. expandBB, for example, can be used to plot an sf object in context: it takes a numeric vector of length four that expands the bounding box of the plot relative to zero in the following order: bottom, left, top, right. This is used to plot Kenya in the context of Africa, in the following code:

```{r}
kenya = world[world$name_long == "Kenya", ]
plot(st_geometry(kenya), expandBB = c(3, 2, 3, 2), col = "gray", lwd = 3)
plot(world_africa[0], add = TRUE)
```
