---
title: "Week 3 Exercises"
author:
  - Louise Bernard, Elisabetta Pietrostefani, Capucine Riom and Lorenzo Sileci

date: "10/10/2022"
output: 
  html_document:
    code_folding: hide
    theme: cerulean 
    highlight: tango
---
```{r class.source = "fold-show", setup, include=FALSE}
knitr::opts_knit$set(root.dir = '~/Dropbox/Teaching/GY476/gy476_data_2022_23')
```

### EXERCISE 1 {.tabset .tabset-fade .tabset-pills}  

##### 1. Vectors

* Assign the first 10 elements of the [Fibonacci sequence](https://www.mathsisfun.com/numbers/fibonacci-sequence.html) to a numeric vector called `fibonacci_vector`.
```{r}
fibonacci_vector <- c(0, 1, 1, 2, 3, 5, 8, 13, 21, 34)
```

* Assign the names of the people sitting at your table to a character vector.
```{r}
people_vector <- c("Elisabetta", "Capucine", "Lorenzo")
```

* Inspect the `length` and `class` of your numeric and character vectors.
```{r, results='hide'}
length(fibonacci_vector)
length(people_vector)
class(fibonacci_vector)
class(people_vector)
```

* Construct a numeric vector containing 10 numbers generated at random from the Uniform distribution with interval [0,1] (*Hint:* `runif()`).
```{r}
random_uniform <- runif(10, 0, 1)
```

* Multiply this vector by a scalar.
```{r, results='hide'}
random_uniform*3
```

* Construct a numeric vector by multiplying `fibonacci_vector` by the vector constructed at step 4.
```{r}
new_numeric_vector <- fibonacci_vector*random_uniform
```


##### 2. Matrices

* Construct a 3x3 matrix containing `fibonacci_vector`, the vector of random draws from the uniform distribution, and their multiplication.
```{r}
new_matrix <-  matrix(c(fibonacci_vector, random_uniform, new_numeric_vector), ncol =3)
```

* Convert the matrix to a dataframe (*Hint:* `as.data.frame()`)
```{r}
new_df <-  as.data.frame(new_matrix)
```

* Name the dataframe columns (*Hint:* `dplyr::rename()`)

```{r}
library(tidyverse)

new_df <-  new_df %>%
  dplyr::rename(fibonacci_vector = V1,
                random_uniform = V2,
                new_numeric_vector = V3)
```


##### 3. Data Frames 

* Construct a Data Frame with 5 columns with an ID, City Name, Population, Area and Population density of 3 cities in the UK.  You can use London, Bristol and other cities in the UK.

```{r}

UK_cities = data.frame(
  id = c(1,2,3),
  city_name = c("London", "Bristol", "Liverpool"),
  population = c(8982000, 467099, 864122),
  area = c(1572, 110, 200)
)

UK_cities$pop_density = UK_cities$population/UK_cities$area

# or the tidy way
UK_cities_tidy = UK_cities %>%
  mutate(pop_density = population/area)

# Get the structure of the data frame
str(UK_cities)
# Print the summary
print(summary(UK_cities))
```



### EXERCISE 2 {.tabset .tabset-fade .tabset-pills} 

**Import the districts shapefile into R and read the rows which have "Hackney" as value for DIST_NAME**.

```{r}
library(sf) 
districts <- read_sf("London/Polygons/districts.shp")

districts[districts$DIST_NAME == "Hackney", ]
```

**Select the rows which correspond to Haringey, Hackney and Islington using tidyverse functions**.

```{r}
districts %>%
  filter(DIST_NAME == "Haringey" | DIST_NAME == "Hackney" | DIST_NAME == "Islington")
```

### EXERCISE 3 {.tabset .tabset-fade .tabset-pills} 

**Plot the City of London in the same map but in a different colour. Some guidance on colours can be found e.g. [here](https://bookdown.org/hneth/ds4psy/D-3-apx-colors-basics.html).**

```{r}
# plot london in grey
plot(districts$geometry, col = "lightgrey")

# Add city of London in turquoise to the map
plot(districts[districts$DIST_NAME == "City of London", ]$geometry, # select city of london
     col = "turquoise",
     add = T) # add to the existing map
```


### EXERCISE 4 {.tabset .tabset-fade .tabset-pills} 

**Plot London districts**
**Add the City of London in purple to the map**
**Add Hackney in turquoise to the map.**

```{r}
# plot london in grey
plot(districts$geometry, col = "lightgrey")

# Add city of London in turquoise to the map
plot(districts[districts$DIST_NAME == "City of London", ]$geometry, # select city of london
     col = "purple",
     add = T) # add to the existing map

plot(districts[districts$DIST_NAME == "Hackney", ]$geometry, # select Hackney
     col = "turquoise",
     add = T) # reset
```


### EXERCISE 5 {.tabset .tabset-fade .tabset-pills} 

**Import housesales data and select houses with a sales price less than £500k**
**Select the Hackney from the districts shapefile**
**Try to filter houses within the buffer of Hackney**. *Hint:* Explore spatial join functions in the`sf` manual [here](https://r-spatial.github.io/sf/reference/st_join.html)

```{r}

housesales_clean <- read.csv("London/Tables/housesales.csv") %>% # select the main object
  filter(price < 500000) %>% # remove values above 500,000
  st_as_sf(coords = c(17,18)) %>% # create sf
  st_set_crs(27700) # set crs

hackney = districts %>%
  filter(DIST_NAME == "Hackney")

# buffer
hackney_buffer <- st_buffer(hackney, 1000)
plot(hackney_buffer$geometry)
plot(hackney$geometry, add=T)

# filter
hackney_filter <- st_filter(housesales_clean, hackney)
plot(hackney_filter$geometry, add=T, col="red")
plot(hackney$geometry, add=T)

```

**Produce a graduated map of house sales within the area formed by the districts of Hackney, Haringey and Islington + a buffer of 1km** (*Hint:* `st_union()` helps you dissolve geometries). Make sure the buffer and the Greater London area are also shown on the map.

```{r}
library(viridis)
library(ggplot2)

northeast_london = districts %>%
  filter(DIST_NAME == "Hackney" | DIST_NAME == "Haringey" | DIST_NAME == "Islington") %>%
  st_union()

# buffer
NE_london_buffer <- st_buffer(northeast_london, 1000)

# filter
NE_london_filter <- st_filter(housesales_clean, NE_london_buffer)

# greater London
gr_london = districts %>% st_union()


graduated <- ggplot()+
  geom_sf(data = gr_london)+
  geom_sf(data = NE_london_buffer, inherit.aes = F, aes(fill = "red"), alpha = 0.3)+
  geom_sf(data = NE_london_filter, aes(color = price),  alpha = .75, size = 1.5) +
  scale_color_viridis(label = scales::dollar)+
  scale_fill_manual(name="North London",
                    labels="",
                    values = "red")+
  xlab("") +
  ylab("") +
  theme_minimal()

graduated

```



