---
title: "Heatmaps - Make beautiful maps"
author: "Louise Bernard"
date: "edited 21/10/2021"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lockdowns and change in mobility patterns

Heatmaps are one of the best visualization tools for dense point data. 
Heatmap is an interpolation technique that is useful in determining the density of input features.
Heatmaps are most commonly used to visualize crime data, traffic incidents, housing density etc. 

Here we are going to look at the impact of lockdowns on road accidents in London.

```{r, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if(!require("leaflet")) install.packages("leaflet") 
if(!require("dplyr")) install.packages("dplyr") 
if(!require("sf")) install.packages("sf") 
if(!require("osmdata")) install.packages("osmdata") # open street map background
if(!require("ggplot2")) install.packages("ggplot2") 
if(!require("ggmap")) install.packages("ggmap") # kernel density map
if(!require("RColorBrewer")) install.packages("RColorBrewer")  # nice colour palette
```


## Import accidents data from CSV

The first step of this tutorial is to get the London district and reproject it in WGS84. We are going to use it later to crop the accidents data to London extent. 

``` {r}
getwd()
datafolder <- file.path("../GY476_data_2022_23")  # set data folder

# load london districts
districts <- read_sf(file.path(datafolder,"London/Polygons/districts.shp")) %>%
  st_transform(4326)
```

The second step is really exciting: we are going to download data straight from R. 
This way there is no need to switch from one interface to another, and it will be easier when we have to download a large number of files.

We are going to find the accidents' data from the department for transport road traffic and download it in the London/Tables folder. 

``` {r}
# donwload road accidents data the department of transport
download.file("https://data.dft.gov.uk/road-accidents-safety-data/dft-road-casualty-statistics-accident-2020.csv",
              file.path(datafolder, "london", "Tables", "dft-road-casualty-statistics-accident-2020.csv"))
```

The following step is equally exciting. We are going to clean our data without creating numerous intermediary objects. To this end, we are going to use `dplyr` pipes `%>%` that allows us to chain commands. We just have to remove the first argument of each command (the one that defines the input object) and chain our commands. 

We are going to excecute the commands `read.csv successively` (load csv), `filter` (removing the missing values), `mutate` (create or modify variables), `st_as_sf` (create sf object from coordinates), `st_crop` (crop to London shapefile) and get only one output object. 


``` {r}
# use dyplr pipes to efficiently load, clean the data and transform to sf
accidents <- read.csv(file.path(datafolder, "london", "Tables", "dft-road-casualty-statistics-accident-2020.csv")) %>% # load csv
  filter(!latitude=="NULL") %>% # filter missing coordinates
  filter(!longitude=="NULL") %>%
  mutate(latitude = as.numeric(latitude), # change to numeric
         longitude = as.numeric(longitude),
         month = as.numeric(substr(as.character(date), 4, 5))) %>% # create a month variable
  st_as_sf(coords = c("longitude", "latitude"), # create sf object
                         crs = 4326, 
           remove = FALSE) %>%
  st_crop(districts) # crop to london
```

## Heatmap

It is now time to create our heatmap. First, let's use a nice background from OpenStreetMap. 

``` {r}
# get background map
london_map <- get_map(getbb("London"), maptype = "toner-background", source = "stamen")
```

The next step is to transform my background to a ggmap object. ggmap is an excellent package to create heatmaps. 

``` {r}
# Convert into ggmap object
london_map <- ggmap(london_map, extent="device", legend="none")

```

Heatmaps calculate the density of events for each point on the map.

``` {r}
# Plot a heat map layer: Polygons with fill colors based on
# relative frequency of events
london_map <- london_map + stat_density2d(data=accidents,
 aes(x=longitude, y=latitude, fill=..level.., alpha=..level..), geom="polygon")
london_map

```

A spectral colour gradient would be more appropriate for accidents. 

``` {r}
# Define the spectral colors to fill the density contours
london_map <- london_map + scale_fill_gradientn(colours=rev(brewer.pal(7, "Spectral")))
london_map
```

Finally,let's add a title and create a facet wrap for each month. It is an extremely visual way to represent the drop in mobility (and hence accidents) during the lockdowns. 

``` {r}
# add a title
london_map <- london_map + ggtitle("Road accidents in 2020")

## Plot accident each month
london_map <- london_map + facet_wrap(~month)
london_map 

```

**Task 1:** Repeat **Heatmap** exercise with a city of your choice in England (pick one with enough variations)

# References

https://trucvietle.me/r/tutorial/2017/01/18/spatial-heat-map-plotting-using-r.html

